Phenotypic evolution is more constrained in simpler food webs

### 2019-01-09 manuscript.Rmd
incorporate minor editing suggestions from Jordi.

Global change is simplifying the structure of ecological networks; however, we are currently in a poor position to predict how these simplified communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treat the community context of evolving populations as a ``black box''. <!--In contrast, ecological networks provide an explicit representation of the community context, yet often ignore the role of natural selection.-->Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web complexity shape evolutionary constraints. We conducted a field experiment that manipulated the diversity of insect parasitoids (food-web complexity) that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection in simpler vs. more complex food webs. The adaptive landscape was more neutral in complex food webs because different parasitoid species impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that phenotypic evolution becomes more constrained in simplified food webs. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of the community context [@Bascompte2014; @McCann2012]<!--SENTENCE SEEMS MISPLACED-->. <!--These networks describe the interdependency of populations within an ecological community, providing an effective framework for predicting how the strength and organization of ecological interactions (network structure) affect population and community dynamics [@Bascompte2014; @McCann2012]. Nevertheless, these networks often lack quantitative information on how ecological interactions influence selection on phenotypic traits. Thus, it remains unclear how network structure shapes adaptive landscapes of interacting populations. This represents a major barrier in our current ability to predict how changes in community context will affect phenotypic evolution.--> Here, we integrate ecological networks and adaptive landscapes to understand how community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.
<!-- need to cite more than just Hansen and Houle 2008 here -->

Global change is simplifying the structure of ecological communities, which may influence evolutionary constraints in a number of ways. For example, in a multitrophic community or food web, if different consumers impose directional selection on different traits of a shared resource, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), <!--To examine the effect of network structure on adaptive landscapes, -->we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*; Family Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). <!--Food-web complexity may influence the adaptive landscape in a number of ways. For example, if each parasitoid imposes directional selection on a different trait, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if parasitoids impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), -->We then applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.

@@

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. <!--This curvature matrix describes the indirect effects of selection on genetic constraints within a generation ($\Delta \text{G}=\text{G}(\gamma -\beta \beta^{\text{T}})\text{G}$). For example, -->The sign of diagonal terms of the curvature matrix dictate whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (i.e., less independence). Therefore, we can infer the indirect effects of selection on genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix.

@@

<!--Selective constraints-->
All three phenotypic traits we examined experienced directional selection in the simpler food web, indicating that there is an optimal phenotype that maximizes larval survival (i.e. large chamber diameter, small clutch size, and avoidance of conspecifics). In contrast, we did not observe selective constraints on clutch size and oviposition preference in the more complex food web. This suggests that there are many optimal phenotypes (adaptive peaks), or in this case, a flat adaptive landscape where there are no fitness consequences for phenotypic change in these traits. This also implies that as selective constraints dampen in more complex food webs, then the trajectory of evolution becomes more determined by genetic constraints. This is made clear if we inspect the 'Lande equation', $\Delta z = \text{G} \beta$, where G corresponds to the population's G-matrix. As the components of $\beta$ go to zero, then $\Delta z$ is more influenced by the G-matrix. Interestingly, this corresponds to @Schluter1996's hypothesis that phenotypic evolution often follows 'genetic lines of least resistance'. @Schluter1996 found support for this hypothesis from data on natural populations of several vertebrate species, including threespine sticklebacks, a few species of songbirds, and mice from the genus *Peromyscus*. All of these species occupy intermediate trophic levels and are likely embedded in complex food webs, which is consistent with our suggestion that genetic constraints may have a stronger influence in more complex food webs.

@@

<!-- Community context shaping adaptation to environmental change --> 
There are a growing number of theoretical studies on adaptation to directional environmental change that incorporate species interactions [e.g. @deMazancourt2008; @Johansson2008; @Norberg2012; @Osmond2017PredatorsHelpPrey]. This work has given insight to the mechanisms by which pairwise interactions modify the adaptive potential of species. Our study hints at a novel mechanism, whereby food-web complexity flattens the adaptive landscape, thus facilitating future adaptation by allowing genetic and phenotypic variation to persist. This mechanism only emerges once we move beyond pairwise interactions to consider selection on multiple traits in a community context. However, a study on competition has highlighted that we may expect the opposite effect of species diversity in competitive communities [@deMazancourt2008]. This negative effect of species diversity on adaptation occurs because there is a greater chance that the optimal phenotype is already occupied by a competitor in a more diverse community. More work is needed to examine how the distribution of different interaction types affects adaptation to environmental change in species-rich communities.

>>>>

Global change is simplifying the structure of ecological networks; however, we are currently in a poor position to predict how these simplified communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. <!--In contrast, ecological networks provide an explicit representation of the community context, yet often ignore the role of natural selection.-->Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web complexity shape evolutionary constraints. We conducted a field experiment that manipulated the diversity of insect parasitoids (food-web complexity) that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection in simpler vs. more complex food webs. The adaptive landscape was more neutral in complex food webs because different parasitoid species impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that phenotypic evolution becomes more constrained in simplified food webs. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of the community context [@Bascompte2014; @McCann2012]<!--SENTENCE SEEMS MISPLACED-->. <!--These networks describe the interdependency of populations within an ecological community, providing an effective framework for predicting how the strength and organization of ecological interactions (network structure) affect population and community dynamics [@Bascompte2014; @McCann2012]. Nevertheless, these networks often lack quantitative information on how ecological interactions influence selection on phenotypic traits. Thus, it remains unclear how network structure shapes adaptive landscapes of interacting populations. This represents a major barrier in our current ability to predict how changes in community context will affect phenotypic evolution.--> Here, we integrate ecological networks and adaptive landscapes to understand how the community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how the community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.
<!-- need to cite more than just Hansen and Houle 2008 here -->

Global change is simplifying the structure of ecological communities, which may influence evolutionary constraints in a number of ways. For example, in a multitrophic community or food web, if different consumers impose directional selection on different traits of a shared resource, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities, <!--To examine the effect of network structure on adaptive landscapes, -->we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). <!--Food-web complexity may influence the adaptive landscape in a number of ways. For example, if each parasitoid imposes directional selection on a different trait, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if parasitoids impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), -->We then applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.

@@

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. <!--This curvature matrix describes the indirect effects of selection on genetic constraints within a generation ($\Delta \text{G}=\text{G}(\gamma -\beta \beta^{\text{T}})\text{G}$). For example, -->The sign of diagonal terms of the curvature matrix dictates whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (i.e., less independence). Therefore, we can infer the indirect effects of selection on genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix.

@@

<!--Selective constraints-->
All three phenotypic traits we examined experienced directional selection in the simpler food web, indicating that there is an optimal phenotype that maximizes larval survival (i.e. large chamber diameter, small clutch size, and avoidance of conspecifics). In contrast, we did not observe selective constraints on clutch size and oviposition preference in the more complex food web. This suggests that there are many optimal phenotypes (adaptive peaks), or in this case, a flat adaptive landscape where there are no fitness consequences for phenotypic change in these traits. This also implies that as selective constraints dampen in more complex food webs, then the trajectory of evolution becomes more determined by genetic constraints. This is made clear if we inspect the 'Lande equation', $\Delta z = \text{G} \beta$, where G corresponds to the population's G-matrix. As the components of $\beta$ go to zero, then $\Delta z$ is more influenced by the G-matrix. Interestingly, this aligns with @Schluter1996's hypothesis that phenotypic evolution often follows 'genetic lines of least resistance'. @Schluter1996 found support for this hypothesis from data on natural populations of several vertebrate species, including threespine sticklebacks, a few species of songbirds, and mice from the genus *Peromyscus*. All of these species occupy intermediate trophic levels and are likely embedded in complex food webs, which is consistent with our suggestion that genetic constraints may have a stronger influence in more complex food webs.

@@

<!-- Community context shaping adaptation to environmental change --> 
There is a growing number of theoretical studies on adaptation to directional environmental change that incorporate species interactions [e.g. @deMazancourt2008; @Johansson2008; @Norberg2012; @Osmond2017PredatorsHelpPrey]. This work has given insight to the mechanisms by which pairwise interactions modify the adaptive potential of species. Our study hints at a novel mechanism, whereby food-web complexity flattens the adaptive landscape, thus facilitating future adaptation by allowing genetic and phenotypic variation to persist. This mechanism only emerges once we move beyond pairwise interactions to consider selection on multiple traits in a community context. However, a study on competition has highlighted that we may expect the opposite effect of species diversity in competitive communities [@deMazancourt2008]. This negative effect of species diversity on adaptation occurs because there is a greater chance that the optimal phenotype is already occupied by a competitor in a more diverse community. More work is needed to examine how the distribution of different interaction types affects adaptation to environmental change in species-rich communities.
